\paragraph{Related work}
{\color{red}The challenges of using \LaTeX{} in LLM-driven workflows have been documented from multiple angles. On the generation side, benchmarks reveal a consistent pattern: LLMs struggle with \LaTeX{}'s long-range syntactic dependencies and implicit semantic conventions. The TeXpert benchmark~\cite{kale2025texpert} reports only 15\% accuracy on complex \LaTeX{} tasks, with logical errors---not superficial typos---accounting for 54\% of failures. Similar difficulties arise in modality-crossing tasks: Ling et al.~\cite{ling_table2latex_2025} found that reinforcement-learning-based table-to-\LaTeX{} conversion suffers from persistent token inefficiency rooted in \LaTeX{}'s verbose syntax, while Xia et al.~\cite{xia2024docgenome} attributed the high edit distances in equation and table conversion (Edit Distance $> 0.21$) to the information entropy of \LaTeX{} source code. These generation-level failures become systemic when \LaTeX{} is embedded in end-to-end scientific pipelines: Lu et al.~\cite{lu2024aiscientist} demonstrated that in a fully automated discovery framework (The AI Scientist), the \LaTeX{} writing stage is a critical bottleneck---the model cannot perceive the rendered PDF, producing overflowing tables and placeholder text that require manual correction. Jain et al.~\cite{jain2026bibby} arrived at a complementary conclusion from the editor side: their Bibby AI system found that raw compiler logs are insufficient for error localization and proposed combining logs with a live AST, effectively acknowledging that \LaTeX{}'s batch feedback model must be augmented with structural representations to support reliable AI assistance. Our work extends this line of reasoning by arguing that such augmentation is inherently limited---a document format designed around explicit tree structure, rather than one retrofitted with AST recovery, can more fundamentally resolve the feedback and semantic transparency problems these systems encounter.

The representation efficiency of \LaTeX{} also poses challenges upstream, during data preparation and model training. Paster et al.~\cite{paster2023openwebmath} documented the substantial engineering effort required to extract and normalize 14.7B tokens of mathematical content from web sources containing heterogeneous \LaTeX{} fragments. Lin et al.~\cite{lin2024accurate} showed that direct LLM processing of documents incurs prohibitive costs, achieving $30\times$ reduction only through semantic hierarchical indexing---a workaround necessitated by the verbosity of the underlying format. That markup structure matters for model learning has been demonstrated more directly: Li et al.~\cite{li2022markuplm} showed with MarkupLM that jointly modeling text and markup tags improves document understanding, while Taylor et al.~\cite{taylor2022galactica} found in training Galactica that \LaTeX{} equations are a first-order component of scientific language whose representational overhead affects model behavior. Taken together, these findings suggest that \LaTeX{}'s syntactic redundancy---where equivalent renderings admit multiple source forms (e.g., \texttt{\textbackslash frac\{a\}\{b\}} vs.\ \texttt{\{a \textbackslash over b\}})---reduces next-token predictability and inflates training costs. In Section~\ref{sec:eff-in-sft}, we provide direct experimental evidence for this hypothesis by comparing fine-tuning convergence on \LaTeX{} versus the lower-entropy \texttt{.tmu} format.

On the systems side, dissatisfaction with \TeX{}'s batch compilation has produced two distinct architectural responses, neither of which fully addresses the problem our work targets. The first is incremental markup compilation: M\"{a}dje~\cite{madje2023typst} and Haug~\cite{haug2022typst} designed Typst with a functional type system and incremental compiler that achieves sub-second preview updates, resolving the latency problem but retaining a source-editing paradigm that still separates authoring from visual output. The second is interactive augmentation of existing \LaTeX{} workflows: Gobert and Beaudouin-Lafon~\cite{gobert2022ilatex} proposed i-\LaTeX{}, which overlays interactive ``transitional'' widgets onto a code editor to bridge source and rendered output---an approach that alleviates feedback delays without removing the underlying batch compilation dependency. Both strategies leave the core tension unresolved: the document's authoritative representation remains either linear markup (Typst) or \TeX{} source (i-\LaTeX{}), rather than an explicit semantic tree. The structured WYSIWYG approach pioneered by Van der Hoeven~\cite{van_der_hoeven_gnu_2001} with GNU \TeX{}macs took a more radical position, demonstrating that high-quality mathematical typesetting can be achieved within a tree-based editor that maintains structure as its primary representation. Empirical evidence supports the case for such alternatives: Knauff and Nejasmic~\cite{knauff2014efficiency} showed in controlled experiments that \LaTeX{} users can be slower and produce more formatting errors than Word users, while Tan and Rigger~\cite{tan2024inconsistencies} systematically documented visual inconsistencies across \TeX{} engines and TeX Live versions, substantiating the fragility of the ecosystem under real-world version drift. Our work builds on the \TeX{}macs lineage through Mogan STEM (Section~\ref{sec:mogan}), and contributes new evidence that the tree-structured \texttt{.tmu} format not only improves editing responsiveness but also yields measurable advantages for LLM tasks---a dimension absent from prior structured-editor evaluations.}
Paster et al.~\cite{paster2023openwebmath} attach the importance to mathematical content for LLM training. They constructed a 14.7B token dataset from web sources and demonstrated the complexity of extracting and processing \LaTeX{} content from diverse formats.

Kale and Nadadur~\cite{kale2025texpert} demonstrate that even state-of-the-art LLMs struggle with \LaTeX{} code generation. The LLMs achieved only 15\% accuracy on complex tasks, with logical errors accounting for 54\% of failures.

Xia et al.~\cite{xia2024docgenome} performed large-scale benchmarks to reveal that Equation-to-\LaTeX{} and Table-to-\LaTeX{} conversion tasks remain challenging even with extensive training data (Edit Distance > 0.21). The reason is the high information entropy of \LaTeX{} source code.

Lin et al.~\cite{lin2024accurate} highlight the high computational cost of applying LLMs directly to document processing. Their approach achieves up to $30\times$ cost reduction through semantic hierarchical indexing, highlighting the need for more efficient document representations.

Evidence from neural code modeling further supports the importance of structured representations. TreeDiff~\cite{zeng2026treediffastguidedcodegeneration} demonstrated that AST-guided span masking significantly outperforms token-level random masking for diffusion-based code generation, particularly for longer sequences (36.59\% vs.\ 33.54\% pass@1 on HumanEval with 1024-token prompts). This result suggests that hierarchical representations enable models to better capture long-range dependencies. Similarly, our tree-structured \texttt{.tmu} format provides a more information-dense and structurally explicit representation than linear \texttt{.tex}, enabling more efficient learning.

Lyn and Graham~\cite{lyn2025translatex} identify an ``execution illusion'' where LLMs produce linguistically fluent but unexecutable \LaTeX code for scientific formatting. They introduce TransLaTeX, a reasoning-and-control framework with compiler-level verification, and SafeFormatBench for executable \LaTeX generation and template conversion.

Gardner et al.~\cite{gardner2025neuralatex} implement a deep learning library entirely in pure \LaTeX; when the document is compiled, the \LaTeX engine trains the network and generates figures. Their paper took 48 hours to compile, illustrating the extreme runtime cost of \LaTeX{} when used as a programmable substrate.
